Choropleth maps using R's base graphics
========================================================

This example shows how to create a basic choropleth map in R
as an add-on to a tutorial on R for spatial data visualisation:
https://github.com/geocomPP/sdvwR

In this tutorial we focus on ggplot2 for making attractive visualisations. 
However, R's base graphics also allow for map creation and 
choropleth maps. 

As before, the first stage is to load the data.

```{r}
load("data/lndStC.RData")
```

There are many colour schemes available and they can be customised manually. 
This can be time consuming, however, so it is recommended to use a colour package
that can create colour schemes automatically. 
Load the package and see what are available:

```{r}
library(RColorBrewer)
brewer.pal.info
```

Now we create the breaks and plot the result:

```{r Choropleth map of number of transport nodes in London boroughs}
cols <- brewer.pal(4, "Greens")
brks <- c(4, 12, 19, 29, 55)
cut(lndStC$NUMBER, brks)
gs <- cols[findInterval(lndStC$NUMBER, vec = brks)]
# png(filename="figure/nStations.png", width = 600, height= 550) # add to save plot
plot(lndStC, col = gs)
legend("topleft", legend = levels(cut(lndStC$NUMBER, brks)), 
       fill = cols, title="N. stations")
# dev.off() # used to stop the png graphics driver
```

