Harmongraphs
========================================================

```{r, echo=FALSE}
library(scales)
htheme <- theme_classic() + theme(
        axis.line = element_line(size = 3, colour = "white", linetype = "dotted")) 
notheme <- htheme + theme(axis.text = element_blank(),
                          axis.ticks = element_blank(),
                          axis.title = element_blank())
```


Harmonographs are made when two or more
pendulums are joined together. The finely 
spaced lines results from the slowdown in the pendulum's 
swing, such that each successive curve is drawn slightly
closer to the centre. 

The basic equation of a pendulum swing over time (t)
is:

$$x = A sin(tf+p)e^{-dt}$$

where *x* is the axis of motion, $A$ is the size of the initial swing, $f$
is the frequency of swing and $p$ is the phase, representing when the swing starts.
$d$ controls the rate of decay in the exponential decay ($e^{-dt}$) bit.
The movement of a pendulum is simply variations on the above equation. 
In R code, this can be described as follows:

```{r}
# set parameters
A = 1
f = 5
p = 0
d = 0.4
t = seq(from = 0, to = 10, by = 0.01)

x = A * sin(t * f + p) * exp(-d * t)

qplot(x = t, y = x, geom="line") + htheme
```

When another axis is defined, you can plot against x and y for 
the duration of the harmonograph's movement:

```{r}
t = seq(from = 0, to = 30, by = 0.01); 
x = A * sin(t * f + p) * exp(-d * t)
x2 = 2 * A * sin(t * 10 + p) * exp(-0.2 * t)

qplot(x, x2, geom="path") + notheme

qplot(x, x2, geom="path", size = I(t)) + geom_path(linejoin="mitre") + notheme
```

More complex harmonographs use two or more interconnected pendulums.

```{r}
x = A * sin(t * f + p) * exp(-0.1 * t) + 
  2 * A * sin(t * 10 + p) * exp(-0.2 * t)
y = 0.5 * A * sin(t * 20 + p) * exp(-0.1 * t) + 
  2 * A * sin(t * 10 + 1) * exp(-0.2 * t)
qplot(x, y, geom="path") + notheme
```

```{r}
capt <-   paste(
'The digital snail:
x = A * sin(t * f + p) * exp(-0.1 * t) + 
2 * A * sin(t * 10 + p) * exp(-0.2 * t);
y = 0.5 * A * sin(t * 20 + p) * exp(-0.1 * t) + 
2 * A * sin(t * 10 + 1) * exp(-0.2 * t)
qplot(x, y, geom="path") + notheme
Made with free software ggplot2. 
Happy Christmas!')

qplot(x, y, geom="path") + notheme + geom_text(aes(x = -1.2, y = 1.6, label =capt), size = 4, ) + theme(panel.border = element_rect(colour = "black", fill=NA))

ggsave("snail.pdf")
```


